---
title: "ST558 HW 9 - Modeling Practice (HW 8 cont.)"
format: html
editor: visual
---

# HW 8 - Basic Modeling Practice
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| warning: FALSE
library(tidyverse)
library(readr)
library(lubridate)
library(janitor)
library(corrplot)
library(tidymodels)
```

## Basic Modeling Practice

### Reading Data

Today we will do some structured practice with fitting linear models in R. The data set we will be using is about bike sharing rentals in Seoul, South Korea. Let's begin by reading in this data.

```{r}
#| results: 'hide'
bike_data <- read_csv("SeoulBikeData.csv", locale = locale(encoding="latin1")) # Locale solves problems with special characters in other languages.
```

### EDA

#### Data Processing

Before we go fitting any models, we will conduct some basic data cleaning to prepare our data for usage. First, we will check for missing values and verify data types to make sure the columns were read in correctly.

```{r}
#| results: 'hide'
# Check structure of data
str(bike_data)

# Check for missing values
colSums(is.na(bike_data)) 
```

We can see that none of the columns have missing values! However, we should note that the `Date` column has been read in as a character type, so we will need to process this column further. The categorical variables `Seasons`, `Holiday`, and `Functioning Day` have also been read in as character type, so we will convert them to factor type for later analysis. The `Date` column is currently stored as a character string with the format day/month/year. We will convert it into a usable date object using the `lubridate` library.

```{r Date Conversion}
bike_data <- bike_data |>
  mutate(Date = dmy(Date))
```

Next, we will make sure that the categorical variables are read as factors.

```{r Characters to Factors}
bike_data <- bike_data |>
  mutate(Seasons = as.factor(Seasons),
         Holiday = as.factor(Holiday),
         `Functioning Day` = as.factor(`Functioning Day`))
```

Finally, we will rename all the variables to have more "user-friendly" names. For example, a column name like "Functioning Day" is not currently user-friendly because of the space between the words.

```{r}
bike_data <- bike_data |>
  clean_names() # Replaces all spaces with '_' and makes all letters lowercase.

names(bike_data)
```

#### Exploratory Data Analysis

Now that the bike data has been cleaned up, we can begin some basic exploratory data analysis (EDA). Let's start by creating some summary statistics as they relate to the bike rental count. We will view these statistics across some categorical variables as well.

```{r}
#| results: 'hide'
# Summary of rented_bike_count variable.
summary(bike_data$rented_bike_count)

# Find unique values for each categorical variable
unique(bike_data$seasons)
unique(bike_data$holiday)
unique(bike_data$functioning_day)
```

Since the `functioning_day` variable only has Yes or No values, this means that the bike sharing system was either operable or inoperable that hour. If the system was inoperable, then we get no useful information about bike rentals for that hour. In this case, we will remove observations that show the system was not operating.

```{r}
#| results: 'hide'
table(bike_data$functioning_day) # 295 "No " observations

bike_data <- bike_data |>
  filter(functioning_day == "Yes")
```

The data is currently given as hourly observations. We will summarize across the hours so that each day has one observation associated with it. We will also group the data by `date`, `seasons`, and `holiday`. We will find the sum of the `rented_bike_count`, `rainfall`, and `snowfall` variables. Finally, we will included the mean of all the weather related variables.

```{r}
daily_bike_data <- bike_data %>%
  group_by(date, seasons, holiday) %>%
  summarise(
    total_bike_count = sum(rented_bike_count),
    avg_temp_c = mean(temperature_c, na.rm = TRUE),
    avg_humidity_pct = mean(humidity_percent, na.rm = TRUE),
    avg_windspeed_m_s = mean(wind_speed_m_s, na.rm = TRUE),
    avg_visibility_10m = mean(visibility_10m, na.rm = TRUE),
    avg_dew_point_temp_c = mean(dew_point_temperature_c, na.rm = TRUE),
    avg_solar_radiation_mj_m2 = mean(solar_radiation_mj_m2, na.rm = TRUE),
    total_rainfall_mm = sum(rainfall_mm, na.rm = TRUE),
    total_snowfall_cm = sum(snowfall_cm, na.rm = TRUE)
  )
```

Now, we have a `daily_bike_data` tibble, which will be our new data that we will analyze! Let's recreate our basic summary statistics and then create some plots to explore variable relationships. We can also view a correlation matrix between our numeric variables to see if the data shows any significant correlations.

```{r}
# Summary of total_bike_count variable on a given day of the year.
summary(daily_bike_data$total_bike_count)
```

```{r Corr Plot}
num_vars <- daily_bike_data[ , 4:12]

corrplot(cor(num_vars))
```

By looking at the correlation plot above, we can see that the total bike rental count has a strong, positive correlation with the average temperature and solar radiation. This makes sense because we can easily imagine more bikes being rented on a warm, sunny day rather than a cold, cloudy day.

#### Split the Data

Now, we will move on to splitting our `daily_bike_data` into training and testing sets (75/25 split).

```{r Splitting}
set.seed(75)

bike_split <- initial_split(daily_bike_data, prop = 0.75, strata = seasons)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
```

We will also create a 10 fold cross-validation (CV) split on the training set.

```{r 10 Fold CV}
bike_folds <- vfold_cv(bike_train, v = 10)
```

#### Fitting MLR Models

Our bike data is now ready for fitting linear regression models! First, let’s create some recipes.

For the 1st recipe, let’s ignore the `date` variable for modeling and instead use it to create a weekday/weekend (factor) variable.

```{r Recipe 1}
#| results: 'hide'
bike_recipe_1 <- recipe(total_bike_count ~ ., data = bike_train) |>
  step_log(total_bike_count, base = exp(1)) |> # Added log for scaling purposes.
  
  update_role(date, new_role = "ID") |> # Give 'date' an ID to keep it, but not for modeling.
  step_date(date, features = "dow", label = TRUE) |> # Pull out week days from date.
  
  # Create new weekday/weekend factor variable.
  step_mutate(
    day_type = factor(dplyr::if_else(date_dow %in% c("Sat", "Sun"), "weekend", "weekday"))) |>
  
  step_rm(date_dow) |> # Remove intermediate day of week variable.
  step_dummy(all_nominal_predictors()) |> # Dummy variables for cat vars.
  step_normalize(all_numeric_predictors()) |> # Standardize num vars.
  step_zv(all_predictors()) # For zero-variance predictor that has no information within the column.

# Check summary
summary(bike_recipe_1)
```

For the 2nd recipe, we will repeat the steps above, as in recipe 1, but now we will add in **interactions** between seasons and holiday, seasons and temp, temp and rainfall.

```{r Recipe 2}
#| results: 'hide'
bike_recipe_2 <- recipe(total_bike_count ~ ., data = bike_train) |>
  step_log(total_bike_count, base = exp(1)) |> # Added log for scaling purposes.
  
  update_role(date, new_role = "ID") |> # Give 'date' an ID
  step_date(date, features = "dow", label = TRUE) |> # Pull out week days
  
  # Create weekday/weekend variable
  step_mutate(
    day_type = factor(if_else(date_dow %in% c("Sat", "Sun"), "weekend", "weekday"))) |>
  
  step_rm(date_dow) |> # Remove intermediate variable
  
# Interactions
  step_interact( ~ starts_with("seasons_"):starts_with("holiday_")) |> # seasons x holiday
  step_interact( ~ starts_with("seasons_"):starts_with("avg_temp_c")) |> # seasons x temp
  step_interact( ~ starts_with("avg_temp_c"):starts_with("total_rainfall_mm")) |> # temp x rainfall
  
  step_dummy(all_nominal_predictors()) |> # Dummy variables
  step_normalize(all_numeric_predictors()) |> # Standardize
  step_zv(all_predictors())

summary(bike_recipe_2)
```

For the 3rd and final recipe, we will repeat the steps above, as in recipe 2, but now we will add in **quadratic terms** for each numeric predictor.

```{r Recipe 3}
#| results: 'hide'
bike_recipe_3 <- recipe(total_bike_count ~ ., data = bike_train) |>
  step_log(total_bike_count, base = exp(1)) |> # Added log for scaling purposes.
  
  update_role(date, new_role = "ID") |> # Give 'date' an ID
  step_date(date, features = "dow", label = TRUE) |> # Pull out week days
  
  # Create weekday/weekend variable
  step_mutate(
    day_type = factor(if_else(date_dow %in% c("Sat", "Sun"), "weekend", "weekday"))) |>
  
  step_rm(date_dow) |> # Remove intermediate variable
  
  step_poly(all_numeric_predictors(), degree = 2) |> # ADD IN QUADRATIC TERMS
  
# Interactions
  step_interact( ~ starts_with("seasons_"):starts_with("holiday_")) |> # seasons x holiday
  step_interact( ~ starts_with("seasons_"):starts_with("avg_temp_c")) |> # seasons x temp
  step_interact( ~ starts_with("avg_temp_c"):starts_with("total_rainfall_mm")) |> # temp x rainfall
  
  step_dummy(all_nominal_predictors()) |> # Dummy variables
  step_normalize(all_numeric_predictors()) |> # Standardize
  step_zv(all_predictors())

summary(bike_recipe_3)
```

Finally, we will set up our linear model fit to use the “lm” engine. We will fit the models using 10 fold CV, then consider the training set CV error to choose a best model.

```{r Additional Packages}
library(parsnip)
library(workflows)
library(broom)
```

Here, we will start by building a model specification. As stated earlier, we will be using a *linear model* on our bike data.

```{r Model Specification}
lm_mod <- 
  linear_reg() |> 
  set_engine("lm")
```

Then, we can create some model workflows, which pair models and recipes together. We will need to do this for all three recipes.

```{r Workflows}
bike_wf1 <- 
  workflow() |>
  add_model(lm_mod) |>
  add_recipe(bike_recipe_1)

bike_wf2 <- 
  workflow() |>
  add_model(lm_mod) |>
  add_recipe(bike_recipe_2)

bike_wf3 <- 
  workflow() |>
  add_model(lm_mod) |>
  add_recipe(bike_recipe_3)
```

Next, we will fit the models using 10 fold CV.

```{r}
set.seed(75)

cv_wf1 <- bike_wf1 |>
  fit_resamples(bike_folds)

cv_wf2 <- bike_wf2 |>
  fit_resamples(bike_folds)

cv_wf3 <- bike_wf3 |>
  fit_resamples(bike_folds)
```

Now, we can look at the training set CV error to choose our best model.

```{r Observe RMSE}
collect_metrics(cv_wf1)
collect_metrics(cv_wf2)
collect_metrics(cv_wf3) # Lowest RMSE
```

Since our third model seems to have the lowest RMSE, we will fit this model to the entire training data set. Then, we will compute the RMSE metric on the test set.

```{r}
#| results: 'hide'
best_fit <- bike_wf3 |>
  last_fit(bike_split, metrics = metric_set(rmse, mae))

best_fit |> collect_metrics()
```

To end, we will obtain the final model (fit on the entire training set) coefficient table.

```{r}
#| results: 'hide'
final_model_coeffs <- best_fit |>
  extract_fit_parsnip() |>
  tidy()

final_model_coeffs
```



# HW 9 - Modeling Practice
      
For the overall best model, fit it to the entire data set!
```{r}
#| warning: FALSE
library(tune)
library(glmnet)
library(dials)
library(baguette)
library(ranger)
library(rpart.plot)
library(vip)
```


In this section, we will be working off of the analysis we previously did in homework 8. Using the first recipe we came up with, we will add several new models. We will begin by adding a **LASSO model** to fit on our bike data. In this case, we will still be using linear regression, but we will be using the "glmnet" engine instead of "lm".
```{r LASSO spec}
LASSO_recipe <- bike_recipe_1

LASSO_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
```

Next, we will create our workflow.
```{r LASSO wkf}
LASSO_wkf <- workflow() |>
  add_recipe(LASSO_recipe) |>
  add_model(LASSO_spec)
LASSO_wkf
```

Below, we will use the `tune_grid()` function. This function allows us to fit the model to CV folds but specify the set of tuning parameters to consider.
```{r LASSO grid}
#| warning: FALSE
set.seed(75)

LASSO_grid <- LASSO_wkf |>
  tune_grid(resamples = bike_folds,
            grid = grid_regular(penalty(), levels = 150)) 
```

Now, we want these metrics computed across the folds for each of our 150 values of the tuning parameter.
```{r LASSO metric}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse")
```

We will find the best model from our LASSO models and fit it to the entire training data set and see how it predicts on the test set.
```{r LASSO finalize & fit}
best_LASSO <- LASSO_grid |> 
  select_best(metric = "rmse")

LASSO_final <- LASSO_wkf |> 
  finalize_workflow(best_LASSO) |>
  last_fit(bike_split, metrics = metric_set(rmse, mae))
```


## Regression Tree Model

We will repeat the same steps, but instead use a **regression tree model**. First, we’ll define our recipe and model specifications.
```{r Tree spec}
tree_recipe <- bike_recipe_1

tree_mod <- decision_tree(tree_depth = tune(),
                          min_n = 20,
                          cost_complexity = tune()) |>
  set_engine("rpart") |>
  set_mode("regression")
```

Next, we will create our workflow.
```{r Tree wkf}
tree_wkf <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(tree_mod)
tree_wkf
```

Then, we will fit the workflow to our CV folds.
```{r Tree fit wkf to folds}
set.seed(75)

tree_grid <- grid_regular(cost_complexity(),
                          tree_depth(),
                          levels = c(10, 5))

tree_fits <- tree_wkf |> 
  tune_grid(resamples = bike_folds,
            grid = tree_grid)
tree_fits
```

Now that the model has been tuned, we can collect metrics to see how the models performed.
```{r Tree metrics}
tree_fits |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean)

best_tree <- tree_fits |> 
  select_best(metric = "rmse")
best_tree
```

Finally, we will finalize our workflow and fit it to the entire training data set.
```{r Tree finalize & fit}
tree_final <- tree_wkf |>
  finalize_workflow(best_tree) |>
  last_fit(bike_split, metrics = metric_set(rmse, mae))

tree_final |>
  collect_metrics()
```


## Bagged Tree Model

We will repeat the same steps again, but instead use a **bagged tree model**. First, we’ll define our recipe, model specifications, and create the workflow.
```{r Bag spec & wkf}
bag_recipe <- bike_recipe_1

# Model specification
bag_spec <- bag_tree(tree_depth = 5, min_n = 10, cost_complexity = tune()) |>
  set_engine("rpart") |>
  set_mode("regression")

# Workflow
bag_wkf <- workflow() |>
  add_recipe(bag_recipe) |>
  add_model(bag_spec)
bag_wkf
```

Next, we will move on to fitting the workflow to our CV folds and checking our metrics across the folds.
```{r Bag fit to folds & check metrics}
set.seed(75)

bag_grid <- grid_regular(cost_complexity(), levels = 15)

bag_fit <- bag_wkf |>
  tune_grid(resamples = bike_folds,
            grid = bag_grid)

# Collect metrics
bag_fit |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean)

# Check metric performance
best_bag <- bag_fit |> 
  select_best(metric = "rmse")
best_bag
```

Finally, we will finalize and fit the model onto the training data set.
```{r Bag finalize & fit}
bag_final <- bag_wkf |>
  finalize_workflow(best_bag) |>
  last_fit(bike_split, metrics = metric_set(rmse, mae))

bag_final |>
  collect_metrics()
```


## Random Forest Model

We will repeat the same steps again, but instead use a **random forest model**. First, we’ll define our recipe, model specifications, and create the workflow.
```{r RF spec & wkf}
rf_recipe <- bike_recipe_1

# Model specification
rf_spec <- rand_forest(mtry = tune()) |>
  set_engine("ranger", importance = "permutation") |> # Added level for importance
  set_mode("regression")

# Workflow
rf_wkf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
rf_wkf
```

Next, we will move on to fitting the workflow to our CV folds and checking our metrics across the folds.
```{r RF fit to folds & check metrics}
set.seed(75)

rf_fit <- rf_wkf |>
  tune_grid(resamples = bike_folds,
            grid = 7)

# Collect metrics
rf_fit |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean)

# Check metric performance
best_rf <- rf_fit |> 
  select_best(metric = "rmse")
best_rf
```

Finally, we will finalize and fit the model onto the training data set.
```{r RF finalize & fit}
rf_final <- rf_wkf |>
  finalize_workflow(best_rf) |>
  last_fit(bike_split, metrics = metric_set(rmse, mae))

rf_final |>
  collect_metrics()
```


## Comparing All Final Models

In this last section, we will compare all final models on the test set using both RMSE and MAE (mean absolute error).
```{r}
all_metrics <- rbind(
  best_fit |>
    collect_metrics() |>
    mutate(Model = "MLR", .before = ".metric"),
  LASSO_final |>
    collect_metrics() |>
    mutate(Model = "LASSO", .before = ".metric"),
  tree_final |>
    collect_metrics() |>
    mutate(Model = "REGRESSION_TREE", .before = ".metric"),
  bag_final |>
    collect_metrics() |>
    mutate(Model = "BAGGED_TREE", .before = ".metric"),
  rf_final |>
    collect_metrics() |>
    mutate(Model = "RANDOM_FOREST", .before = ".metric")
)

all_metrics |>
  filter(.metric %in% c("rmse", "mae")) |> 
  arrange(.estimate)
```

According to RMSE, we can see that the best model for our data is the Random Forest Model!

## Final Model Fits & Summaries

For the LASSO and MLR models, we will report the final coefficient tables.
```{r coefficient tables}
tidy(best_fit |> extract_fit_parsnip()) |> arrange(desc(abs(estimate))) # MLR
tidy(LASSO_final |> extract_fit_parsnip()) |> arrange(desc(abs(estimate))) # LASSO
```

For the regression tree model, let's create a plot of the final fit.
```{r Tree final fit plot}
tree_obj <- tree_final |> 
   extract_workflow() |> 
   extract_fit_engine()
 
 # Plot
 rpart.plot(tree_obj, 
            type = 2, 
            fallen.leaves = TRUE, 
            main = "Regression Tree Final Fit")
```

For the bagged tree and random forest models, we can produce variable importance plots.
```{r Bagged variable importance}
# Extract final model fit
bag_final_model <- bag_final |> extract_workflow() |> extract_fit_engine()

bag_final_model$imp |>
  mutate(term = factor(term, levels = term)) |> 
  ggplot(aes(x = term, y = value)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Bagged Tree Variable Importance",
       x = "Variables",
       y = "Importance")
```

```{r RF variable importance}
# Extract final model fit
rf_final_model <- rf_final |> extract_workflow() |> extract_fit_engine()

rf_import <- rf_final_model$variable.importance |> 
  sort(decreasing = TRUE)

# Tibble for plotting
rf_import_df <- tibble(
  variable = names(rf_import),
  importance = rf_import
)

rf_import_df |> 
  ggplot(aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(title = "Random Forest Variable Importance",
       x = "Variables",
       y = "Importance")
```
In both the bagged tree and random forest variable importance plots above, we can see that the most important predictor for the number of bike rentals is the average temperature (c) outside.

To end, we will fit the overall best model to the entire data set. This will be our Random Forest Model.
```{r}
final_rf_wkf <- rf_wkf |>
  finalize_workflow(best_rf)

final_rf_fit <- final_rf_wkf |>
  fit(data = daily_bike_data)

final_rf_model <- final_rf_fit |>
  extract_fit_engine()

final_rf_model
```
Let's break down some of the key points of our random forest model performance. The final random forest model is a regression model that was trained on the full bike data set using 500 trees. This model is able to predict daily bike rental counts pretty accurately, with about 90.7% of the variance in predictions able to be explained and a low OOB (out of bag) prediction error (MSE) of about 0.055. As seen previously in the variable importance plot for this model, the average temperature (c) seems to be the most important variable in the data set when predicting daily bike rental counts.


